A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-14 text/manuscript.Rmd

In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
the highest pathway enrichment scores of the three major categories, though
there is a significant number of tumors spatially distributed between cluster
groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
pre-replicative complex," "metabolism of polyamines," "degradation of
$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
with aspects of hedgehog signaling. To determine once again whether the genes
driving pathway enrichment were commonly shared among these pathways we plotted
and performed unsupervised clustering on tumors with an individual mutation
matrix using every gene in each of the ten pathways, grouped by shared pathway
membership (**figure 4**). Surprisingly, this class of tumors was not
predominated by 1 or more highly mutated genes across all cancers. Instead,
there is a whole set of genes in the proteosomal pathway whose activity
contributes to all the aforementioned pathways, and the majority of tumors in
these classes have a single mutation in one of these genes. Nonetheless, a
significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
proteasomal mutations but instead harbor mutations in genes representing a
majority of the same pathways, suggesting that most if not all of the pathways
are individually important and that proteasomal mutations may be sufficient to
drive the net effect of the other mutations in this cluster. We annotated the
tumors in this class according to tissue of origin to determine whether
proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
indeed we could observe some clustering. These data could explain how
tissue-specific gene mutation signatures might result in general cancer
phenotypes by converging at the pathway level.
```{r child='figure_4.Rmd'}
```
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey). The genes are organized according to which of the top distinguishing pathways for this class that they effect. Some genes affect multiple pathways, as indicated by the bars at left. The highlighted genes at right are all proteasomal subunits.</p>

>>>>

In contrast to clusters 9-12, the group represented by clusters 6 & 7 exhibited
the highest pathway enrichment scores of the three major categories, though
there is a significant number of tumors spatially distributed between cluster
groupings 9-12 and 6-7 on the UMAP projection that have pathway characteristics
of both. The top ten distinguishing pathways of class 6/7 tumors were "cellular
response to hypoxia," "regulation of mitotic cell cycle," "assembly of the
pre-replicative complex," "metabolism of polyamines," "degradation of
$\beta$-catenin by the destruction complex", "tumor necrosis factor receptor 2
non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0) binds and
destabilizes RNA", "regulation of apoptosis" and 3 other pathways associated
with aspects of hedgehog signaling. To determine once again whether the genes
driving pathway enrichment were commonly shared among these pathways we plotted
and performed unsupervised clustering on tumors with an individual mutation
matrix using every gene in each of the ten pathways, grouped by shared pathway
membership (**figure 4**). Surprisingly, this class of tumors was not
predominated by 1 or more highly mutated genes across all cancers. Instead,
genes in the proteosomal pathway whose activity
contributes to all the aforementioned pathways and the majority of tumors in
these classes have a single mutation in one of these genes. Nonetheless, a
significant fraction <mark>(~1/3 <-- need numbers!)</mark> do not have
proteasomal mutations but instead harbor mutations in genes representing a
majority of the same pathways, suggesting that most if not all of the pathways
are individually important and that proteasomal mutations may be sufficient to
drive the net effect of the other mutations in this cluster. We annotated the
tumors in this class according to tissue of origin to determine whether
proteasome mutations have a tissue-of-origin (or cancer-specific) bias, and
indeed we could observe some clustering. These data could explain how
tissue-specific gene mutation signatures might result in general cancer
phenotypes by converging at the pathway level.
```{r child='figure_4.Rmd'}
```
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey). The genes are organized according to which of the top distinguishing pathways for this class that they effect. Some genes affect multiple pathways, as indicated by the bars at left. The highlighted genes at right are all proteasomal subunits.</p>

The proteasome plays key roles in the transduction of signals from the p53, WNT
and NFkB pathways and it and proteasomal inhibitors have been approved for
clinical use in some patient populations [@proteasome_review_Jang]. To determine
whether this class of proteins is broadly sensitive to cancer-promoting
mutations, we tabulated each gene as prognostic or not in the human protein
expression atlas [@protein_atlas] and found 49 positive and 10 negative
associations ( _i.e._ higher expression associated with unfavorable outcomes is
"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 4b**).
Second we looked at survival in mutations results results results (**figure 4c**). Third, we checked the mutation classes and found the overwhelming
majority (x out of y) are missense _vs._ various other deletarious classes (see
**figure 4d**), suggesting that, similar to other oncogenic driver classes such
as RAS and PI3K, proteasomal subunit mutations may result in specific
dysregulation of the proteasomal complex activities that target cancer-promoting
pathways. Taken together, our data suggest that mutations in proteasomal
subunits are more widespread in cancer than previously appreciated and may
represent a targetable category of tumors.
